Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-08-02 jso.qmd
a proposition

Squintability has been a useful concept. Here, it is defined
mathematically and two approaches to compute this metric numerically,
are described.

@@

We now focus our attention to how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO.

@@

@@

$$\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d) $$

@@

<!-- Definition \ref{def:squintability} is ... -->

@@

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.  
Noticing that $f$ is expected to be a decreasing function, the
squintability metric is proposed as follows:

$$
\varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)\ .
$$ {#eq-squintability}

$-\min_{x}f'(x)$ represents the largest value of the gradient of $-f$
and $\underset{x}{\arg\min}f'(x)$ represents the projection distance at
which this largest gradient is attained. It is expected that these two
values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggest that their product (@eq-squintability) should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

>>>>

Here the formal definition of projection distance and squint angle are given, before the definition of squintability. Two approaches to compute this metric numerically,
are described.

@@

@@

Squintability can be defined as how the index value $f(XA)$ changes with
respect to the projection distance $d(A, A^*)$, over the course of the
JSO:

@@

\begin{equation}
\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d)
\label{eq-squintability}
\end{equation}

@@

It is expected that these two
values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggests that their product (@eq-squintability) should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

From the definition, the following proposition can be derived:

\begin{prop}\label{prop:convex-concave}
Let $g_1(d)$ be a convex function and $g_2(d)$ be a concave function, where $d \in [0, D]$ and $g_1(d), g_2(d) \in [0, 1]$. Assume both functions achieve the same maximum gradient, i.e., $\max g'_1(d_1) = \max g'_2(d_2) = -g_{max}$, it follows that $d_1 < d_2$ and $\varsigma(f_1) < \varsigma(f_2)$.
\end{prop}

@@

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.

<!-- Noticing that $f$ is expected to be a decreasing function, the -->
<!-- squintability metric is proposed as follows: -->

<!-- $$ -->
<!-- \varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)\ . -->
<!-- $$ {#eq-squintability} -->

<!-- $-\min_{x}f'(x)$ represents the largest value of the gradient of $-f$ -->
<!-- and $\underset{x}{\arg\min}f'(x)$ represents the projection distance at -->
<!-- which this largest gradient is attained.  -->

### 2024-08-02 jso.tex
a proposition

\begin{tcolorbox}[enhanced jigsaw, colback=white, opacityback=0, titlerule=0mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, coltitle=black, breakable, left=2mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm]

@@

We now focus our attention to how the index value \(f(XA)\) changes with
respect to the projection distance \(d(A, A^*)\), over the course of the
JSO.

@@

@@

\begin{prop}\label{prop:my_prop}
This is a proposition.
\end{prop}

Proposition \ref{prop:my_prop} is useful.

@@

Noticing that \(f\) is expected to be a decreasing function, the
squintability metric is proposed as follows:

\begin{equation}\phantomsection\label{eq-squintability}{
\varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)\ .
}\end{equation}

\(-\min_{x}f'(x)\) represents the largest value of the gradient of
\(-f\) and \(\underset{x}{\arg\min}f'(x)\) represents the projection
distance at which this largest gradient is attained.

It is expected that these two values should be both high in the case of
high squintability (fast increase in \(f\) early on), and both low in
the case of low squintability (any substantial increase in \(f\) happens
very late, close to the optimal angle). This suggest that their product
(\ref{eq-squintability}) should provide a sensible measure of
squintability.

The multiplicative constant \(4\), which can be deemed arbitrary, does
not change the interpretation of the squintability metric \(\varsigma\);
it is here to adjust the range of values of \(\varsigma\) and simplify
the explicit formula for \(\varsigma\) obtained later on.

To compute the squintability metric (\ref{eq-squintability}) in
practice, several approaches are possible. The first one is to propose a
parametric model for \(f\), and use it to obtain an explicit formula for
\(\varsigma\). Numerical experiments suggest a scaled sigmoid shape as
described below. Define

@@

which depends on three additional parameters, \(\theta_1\),
\(\theta_2\), and \(x_{\max}\), such that \(f(0)=\theta_1\) and
\(f(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric (\ref{eq-squintability})
can be shown to be equal to

@@

Alternatively, one can estimate (\ref{eq-squintability}) in a
nonparametric way, for example by fitting \(f\) using kernel regression,
then numerically estimate the angle at which \(-f'\) attains its highest
value.

@@

\caption{\label{fig-squintability}Illustration of steps to calculate
squintability. For a given projection pursuit problem defined by the
shape to find, data dimension and the index function, 1) sample random
bases given the orthonormality and projection distance contraint, 2)
interpolate the sampled bases to the optimal basis and calculate the
projection distance and the index value for each interpolated basis. 3)
bin the index values by projection distances to obtain the average index
value for each bin, 4) fit the scaled sigmoid function in Equation
\ref{eq-squintability} to the binned index values against projection
distances using non-linear least square, 5) calculate the squintability
measure using Equation \ref{eq-squintability-parametric} with parameters
estimated from the model.}

>>>>

\begin{tcolorbox}[enhanced jigsaw, breakable, arc=.35mm, colframe=quarto-callout-note-color-frame, colback=white, rightrule=.15mm, coltitle=black, toprule=.15mm, toptitle=1mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, leftrule=.75mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, opacitybacktitle=0.6, bottomtitle=1mm, bottomrule=.15mm, left=2mm]

@@

@@

Squintability can be defined as how the index value \(f(XA)\) changes
with respect to the projection distance \(d(A, A^*)\), over the course
of the JSO:

@@

It is expected that these two values should be both high in the case of
high squintability (fast increase in \(f\) early on), and both low in
the case of low squintability (any substantial increase in \(f\) happens
very late, close to the optimal angle). This suggests that their product
(\textbf{?@eq-squintability}) should provide a sensible measure of
squintability. The multiplicative constant \(4\), which can be deemed
arbitrary, does not change the interpretation of the squintability
metric \(\varsigma\); it is here to adjust the range of values of
\(\varsigma\) and simplify the explicit formula for \(\varsigma\)
obtained later on.

From the definition, the following proposition can be derived:

\begin{prop}\label{prop:convex-concave}
Let $g_1(d)$ be a convex function and $g_2(d)$ be a concave function, where $d \in [0, D]$ and $g_1(d), g_2(d) \in [0, 1]$. Assume both functions achieve the same maximum gradient, i.e., $\max g'_1(d_1) = \max g'_2(d_2) = -g_{max}$, it follows that $d_1 < d_2$ and $\varsigma(f_1) < \varsigma(f_2)$.
\end{prop}

@@

To compute the squintability metric (\textbf{?@eq-squintability}) in
practice, several approaches are possible. The first one is to propose a
parametric model for \(f\), and use it to obtain an explicit formula for
\(\varsigma\). Numerical experiments suggest a scaled sigmoid shape as
described below. Define

@@

which depends on three additional parameters, \(\theta_1\),
\(\theta_2\), and \(x_{\max}\), such that \(f(0)=\theta_1\) and
\(f(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric
(\textbf{?@eq-squintability}) can be shown to be equal to

@@

Alternatively, one can estimate (\textbf{?@eq-squintability}) in a
nonparametric way, for example by fitting \(f\) using kernel regression,
then numerically estimate the angle at which \(-f'\) attains its highest
value.

@@

\caption{\label{fig-squintability}Illustration of steps to calculate
squintability. For a given projection pursuit problem defined by the
shape to find, data dimension and the index function, 1) sample random
bases given the orthonormality and projection distance contraint, 2)
interpolate the sampled bases to the optimal basis and calculate the
projection distance and the index value for each interpolated basis. 3)
bin the index values by projection distances to obtain the average index
value for each bin, 4) fit the scaled sigmoid function in Equation
\textbf{?@eq-squintability} to the binned index values against
projection distances using non-linear least square, 5) calculate the
squintability measure using Equation \ref{eq-squintability-parametric}
with parameters estimated from the model.}
